A weakly structured stem for human origins in Africa

### 2022-02-08 paper.tex
fix bmh comment command
complete filling in supp fig captions
edits throughout

\documentclass{article}
\usepackage[super,numbers]{natbib}

@@

Regardless of the model choice for early epochs, inference of human demographic
history for the last 150kya was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135kya, with low to moderate levels of
subsequent gene flow (Table\ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140kya. We conclude that, regardless of model
specification, geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
\sgcomment{Alt formulation: Thus all models suggest that broad geographic patterns of contemporary population structure date back to MIS 5}
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

@@

\sgcomment{I don't love this new structure because it breaks the model description in two parts separated by the bootstrap. Maybe we have to, but I would have preferred a unified description of the models. 
For example, we could start this section by the bootstrap discussion ("In all models, likelihood differences are large and bootstrap confidence interval narrow within each specified model ...  We therefore highlight two models".
Then we describe model 1, describe model 2, and contrast the two. 
 )} \bmh{what do you think of the revised text?}

@@

\begin{figure}[t!]
    \centering
    \includegraphics{figures/fig3.pdf}
    \caption{
        \textbf{Predicted population structure over time.}
        From the best fit models of our two parameterizations (top row:
        continuous migration, bottom row: multiple merger), we computed
        preditions of population structure between contemporaneous populations
        over time ($F_{ST}$ \sgcomment{not sure I want Fst here, vs pi}) and the proportion of drift between pairs of sampled
        populations that aligns with drift between populations in the past
        ($\alpha^2$). \sgcomment{point to supplement, since nonstandard}
    }
    \label{fig:3}
\end{figure}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics{figures/fig4}
    \caption{
        \textbf{Model validation using independent statistics.} (A--C) Using
        our best fit models to the LD and pairwise diversity statistics, we
        simulated expected conditional site-frequency-spectra (cSFS) and
        compared to the observed cSFS from the data. Our inferred models
        provide a good fit to the data, even though these were not used in our
        inference. Across the three populations, ancestral state
        misidentification was consistently inferred to be 1.5--1.7\% for
        intergenic loci (Supp. Methods). (D, E) We used Relate
        \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
        which we used to estimate coalescence rate trajectories and
        and cross-coalescence rates between pairs of populations. While
        coalescence rate distributions are informative statistics about past
        evolutionary processes, interpretation can be hindered by migration and
        population structure, and translating relative cross-coalescence rate
        curves (RCCR) into population divergence times is especially prone
        to misinterpretation. For example, the Mende-Gumuz comparison shows
        a more recent increased RCCR than either population with the British,
        a pattern that is recapitulated under our best-fit model, even though
        the Mende-Gumuz split occurs prior to the Gumuz-British split.
    }
    \label{fig:4}
\end{figure}

>>>>

\documentclass[]{article}
\usepackage[super,numbers]{natbib}

@@

Regardless of the model choice for early epochs, inference of human demographic
history for the last 150kya was remarkably robust. The earliest divergence among
contemporary human populations differentiates the southern African Nama from
other African groups between 110--135kya, with low to moderate levels of
subsequent gene flow (Table\ref{tab:migration-rates}). In none of the
high-likelihood models which we explored did the divergence between Nama and
other populations exceed $\sim$140kya. We conclude that
geographic patterns of contemporary \emph{Homo sapiens} population structure
date back to the Late Middle Stone Age in Africa, likely arising during MIS 5.
%\sgcomment{Alt formulation: Thus all models suggest that broad geographic patterns of contemporary population structure date back to MIS 5}
Although we find evidence for earlier population structure in Africa (see
below), present-day populations cannot be easily mapped onto these more ancient
stem groups as only a small proportion of drift between modern populations can
be attributed to drift between stems (Figures~\ref{fig:3} and
\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).

@@

\sgcomment{I don't love this new structure because it breaks the model description in two parts separated by the bootstrap. Maybe we have to, but I would have preferred a unified description of the models. 
For example, we could start this section by the bootstrap discussion ("In all models, likelihood differences are large and bootstrap confidence interval narrow within each specified model ...  We therefore highlight two models".
Then we describe model 1, describe model 2, and contrast the two. 
 )} \bmhcomment{what do you think of the revised text?}

@@

\begin{figure}[t!]
    \centering
    \includegraphics{figures/fig3.pdf}
    \caption{
        \textbf{Predicted population structure over time.}
        From the best fit models of our two parameterizations (top row:
        continuous migration, bottom row: merger with stem migration), we computed
        predicted differentiation between contemporaneous populations over time
        as $H_{i,j}$, the expected pairwise diversity between two samples, one
        each from populations $i$ and $j$.
        We also computed the proportion of drift between pairs of sampled
        populations that aligns with drift between populations in the past
        ($\alpha^2$, see Section~\ref{sec:f4} in the Supplement).
        While both models inferred population structure to extend deep into the
        past, the ancient structure contributes relatively little to present-day
        differentiation between human populations.
    }
    \label{fig:3}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \includegraphics{figures/fig4}
    \caption{
        \textbf{Model validation using independent statistics.} (A--C) Using
        our best fit models to the LD and pairwise diversity statistics, we
        simulated expected conditional site-frequency-spectra (cSFS) and
        compared to the observed cSFS from the data. Our inferred models
        provide a good fit to the data, even though these were not used in our
        inference. Across the three populations, ancestral state
        misidentification was consistently inferred to be 1.5--1.7\% for
        intergenic loci (Supp. Methods). (D, E) We used Relate
        \citep{Speidel2019-nj} to reconstruct genome-wide gene genealogies,
        which we used to estimate coalescence rate trajectories and
        and cross-coalescence rates between pairs of populations. While
        coalescence rate distributions are informative statistics about past
        evolutionary processes, interpretation can be hindered by migration and
        population structure, and translating relative cross-coalescence rate
        curves (RCCR) into population divergence times is especially prone
        to misinterpretation. For example, the Mende-Gumuz comparison shows
        a more recent increased RCCR than either population with the British,
        a pattern that is recapitulated under our best-fit model, even though
        the Mende-Gumuz split occurs prior to the Gumuz-British split.
    }
    \label{fig:4}
\end{figure}
